Concurrent Programming with CXL

### 2020-06-03 book/paper.tex
Fixed the triangle program
Fixed figures a bit
Added a new introduction

@@

CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
``model checked'' to verify that the program has certain desirable
properties and does not suffer from bugs.  Model checking is
limited---it can only check programs that have ``finite state
spaces.''  But even programs that have infinite state spaces can
often be constrained (for example, by specifying a finite set of
test inputs) so that model checking is possible.  If, for
example, a sorting algorithm is shown to work for any subset of
some set of say, 5, different values, then one can hope that it
will also work for sets that are larger than 5 values or sets that
use more than 5 different values.

@@

\begin{figure}
\begin{center}
\includegraphics[width=2in]{figures/mutex.png}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=2in]{figures/rdwr.png}
\end{center}
\caption{High-level state diagram specification of reader/writer locks with
up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes reading in the critical section; the third is the number of
processes writing in the critical section.}
\label{fig:rdwr}
\end{figure}

@@

@@

Introduced by the famous Dutch computer scientist Edsger Dijkstra,
a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
resources available.  When allocating a resource, a process decrements the
counter using the \texttt{procure} or simply \texttt{P} operation.  The
\texttt{procure} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{vacate} or {V} operation.
Figure~\ref{fig:sema} shows the \texttt{synch} module implementation of
semaphores.

>>>>

\chapter{On Concurrent Programming}

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables and
synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why is this?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential program is mostly \emph{deterministic}.
If you run it twice with the same input, the same things will happen.
Bugs are typically easily reproducible.
\item Each statement and each function can be thought of as happening
\emph{atomically} because there is no other activity interfering with
their execution.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that help with \emph{testing} concurrent code.
The approach is based on \emph{model checking}: instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

Model checking is not a replacement for formal verification.  Formal
verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.  Because model checking tries every possible
execution, this means that the test program needs to be relatively
simple.  In particular, it needs to have a relatively small number of
reachable states.
If the model is too large, it may run for longer than we may care to
wait for or even run out of memory.

So why is this useful?  Consider, for example, a sorting algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all corner cases, including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know necessarily that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.

But model checking can also help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}: predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
those invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof.

So what is CXL?
CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
model checked to verify that the program has certain desirable
properties and does not suffer from bugs.

@@

@@

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/mutex.jpeg}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=2in]{figures/rdwr.jpeg}
\end{center}
\caption{High-level state diagram specification of reader/writer locks with
up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes reading in the critical section; the third is the number of
processes writing in the critical section.}
\label{fig:rdwr}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=2.5in]{figures/pc.jpeg}
\end{center}
\caption{High-level state diagram specification of producers/consumers.}
\label{fig:pc}
\end{figure}

@@

Introduced by the famous Dutch computer scientist Edsger Dijkstra,
a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
resources available.  When allocating a resource, a process decrements the
counter using the \texttt{procure} or simply \texttt{P} operation.  The
\texttt{procure} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{vacate} or {V} operation.
Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
semaphores.
